\documentclass[12pt]{article}
\usepackage{amsmath,amssymb,fullpage,graphics,graphicx,yuniorlab}
\usepackage{problemset,url}

\begin{document}

\title{SNEWS Word Problems}
\author{}
\date{}
\maketitle

\section{Exercises}

\begin{probdesc}{1}{How Big is a Neutrino Detector?}
The sizes of neutrino detectors are often measured in {\em kilotons.}
One kiloton is a thousand tons or two million kilograms, so a
ten-kiloton water-based detector contains twenty million kilograms of
water in its tank (when it is completely filled).  For the following,
it may be useful to know that the density of water is about 1 gram per
cubic centimeter, or 1 kilogram per liter.  Also, according to the way
the metric system was defined, it takes 1000 ($10^3$) liters to make
up one cubic meter.

\end{probdesc}

\begin{probdesc}{2}{How Strong is the Signal?}
\begin{equation}
I(R) = I_0 \left(\frac{R_0}{R}\right)^2.
\end{equation}
\end{probdesc}

\begin{probdesc}{3}{Temperature Scales}
\end{probdesc}

\begin{probdesc}{4}{Stellar Masses}
\end{probdesc}

\begin{probdesc}{5}{Signal Delay}
\end{probdesc}

\begin{probdesc}{6}{Nuclear Burning}
\end{probdesc}

\begin{probdesc}{7}{Half-Lives}
\begin{equation}
N(t) = N_0 \cdot \left(\frac{1}{2}\right)^{t / \tau}.
\end{equation}
\end{probdesc}

\begin{probdesc}{8}{Einstein's Equation}
According to Einstein's Special Theory of Relativity, matter and
energy are interchangeable.  It is possible to convert an amount of
mass into pure energy, which may take the form of light or other
electromagnetic radiation.  The exact rule is given by Einstein's
famous equation,
\begin{equation}
E = mc^2.
\end{equation}
Here, $c$ is the speed of light, roughly $3\times10^8$ meters per
second.
\begin{enumerate}
\item[(a)] How much energy would be released if one kilogram of matter (any
  kind of matter) were turned entirely into energy?

\item[(b)] A key characteristic of {\em antimatter} is that when it comes
  into contact with regular matter, both objects vanish, releasing
  their mass as a flood of energy.  One example of such an {\em
  annihilation} reaction happens when an electron meets its
  antiparticle, called a {\em positron} because it has the same mass
  as an electron but a positive electric charge.  Typically, when an
  electron and a positron annihilate, they produce two photons of
  equal energy.  What is the energy of each photon in MeV?  Convert
  this number to joules.

\item[(c)] According to the quantum theory Max Planck helped found, the
  wavelength of a photon is inversely proportional to its energy.
  Planck's equation says that
  \begin{equation}
    E = \frac{hc}{\lambda}
  \end{equation}
  where $h$ is {\em Planck's constant,} a number which experiments show
  is roughly .  (Don't worry about the units too much
  for this problem.)  What is the wavelength of the photon whose energy
  you calculated in part (b)?  In which part of the electromagnetic
  spectrum does this photon fall---can you see it with the naked eye?

\item[(d)] Astrophysicists estimate that a Type II supernova can release
  joules of energy.  Using Einstein's equation, find how many
  kilograms lighter the supernova remnant must be than the original
  star.  Express this number as a percentage, assuming that the
  original star was twenty times as massive as the Sun.
\end{enumerate}
\end{probdesc}

\section{Useful Information}

\begin{table}[!h]
\begin{tabular}{cccc}
\hline\hline\\
Shape & Characteristic Length & Surface Area & Volume \\
\hline\\
Cube & Side $a$ & $8a^2$ & $a^3$\\
Sphere & Radius $r$ & $4\pi r^2$ & $\frac{4\pi}{3}r^3$\\
Cylinder & Height $h$, Radius $r$ & $2\pi r^2  + 2\pi rh$
& $\pi r^2 h$ \\
\hline\hline\\
\end{tabular}
\caption{Useful geometric formulas for common 3D shapes.}
\end{table}

\begin{table}[!h]
\begin{tabular}{ccccc}
\hline\hline\\
Particle & Mass (kg) & Mass (MeV/$c^2$) & Charge & Spin \\
\hline\\
Electron & $9.109\times10^{-31}$ & $0.511$ & $-1$ & $\frac{1}{2}$\\
Proton & $$ & $$ & $+1$ & $\frac{1}{2}$\\
Neutron & $$ & $$ & $+1$ & $\frac{1}{2}$\\
Photon & $$ & $$ & $0$ & $0$\\
Neutrino & $\approx 0$ & $\approx 0$ & $0$ & $\frac{1}{2}$\\
\hline\hline\\
\end{tabular}
\caption{Basic properties of common particles.  Charges are given
in multiples of the fundamental charge unit, $1.602\times10^{-19}$ coulombs.}
\end{table}


\begin{table}[!h]
\begin{tabular}{cccc}
\hline\hline\\
Fuel & Main Product & $T$ (10$^9$ K) & Duration (yr)\\
\hline\\
H & He & 0.037 & $8.1\times10^6$\\
He & O, C & 0.19 & $1.2\times10^6$\\
C & Ne, Mg & 0.87 & $9.8\times10^2$\\
O & Si, S & 2.0 & $1.3$\\
Si & Fe & 3.3 & 0.031 (11 days)\\
\hline\hline\\
\end{tabular}
\caption{Nuclear ``burning'' stages for a star of 20 solar masses.
Source: ``Massive Star Evolution Through the Ages''
({\tt http://arxiv.org/abs/astro-ph/0211062}).}
\end{table}

\end{document}
